---
title: "Feederwatch: `r params$state` State Report"
author: "Jaxine Wolfe"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: tango
params:
  state: "MD"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```

```{r}
state_birds <- data_clean %>% filter(state %in% params$state)
```

[Project FeederWatch](https://feederwatch.org/explore/raw-dataset-requests/) is a November-April survey of birds that visit backyards, nature centers, community areas, and other locales in North America. Participatory scientists count birds in areas with plantings, habitat, water, or food that attracts birds. The schedule is completely flexible. People count birds as long as they like on days of their choosing, then enter their counts online. This allows anyone to track what is happening to birds around your home and to contribute to a continental data-set of bird distribution and abundance. Project FeederWatch is operated by the Cornell Lab of Ornithology and Birds Canada. 

The data used in this report is from one Feederwatch survey which was posted for [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday/tree/main/data/2023/2023-01-10). The observations for this survey were taken between `r min(data_clean$date)` and `r max(data_clean$date)`.

## Data Summary

```{r}

bird_smry <- state_birds %>% 
  group_by(primary_com_name, sci_name) %>% 
  summarize(Count = sum(how_many)) %>% 
  arrange(desc(Count)) %>% 
  rename(`Scientific Name` = sci_name, 
         `Common Name` = primary_com_name)
  
datatable(bird_smry, 
          options = list(autoWidth = TRUE,
                         searching = TRUE,
                         paging = FALSE,
                         scrollY = 300,
                         # scrollX = 300,
                         scrollCollapse = TRUE), 
          rownames = TRUE)

```

## Data Visualization

### Bird Counts

Plotting bird counts for each species observed and arranging in visually descending order. 

```{r fig.height=8, fig.width=8}
bird_smry %>% 
  # filter(Count > 10) %>%
       plot_ly(x = ~Count, y = ~reorder(`Common Name`, Count), type = "bar") %>% 
       layout(xaxis = list(title = "Number of Individuals"),
              yaxis = list(title = "Common Name"))
```

### Observations Over Time

Let's make a heatmap of observations by species and month.

```{r}
top10 <- bird_smry$`Common Name`[1:10]

state_birds %>% 
  filter(primary_com_name %in% top10) %>% 
  group_by(primary_com_name, Year, Month) %>% 
  summarize(how_many = sum(how_many)) %>%
  mutate(Month = factor(as.character(Month))) %>% 
  # mutate(norm_count = how_many/sum(how_many)) %>% 
  ggplot(aes(Month, primary_com_name, fill = how_many)) +
  geom_tile(color = "white") +
  coord_equal() +
  scale_fill_continuous(name = "Bird Count") +
  ylab("Common Name")
  # guides(fill=guide_legend(title="Bird Count")) # another way to define the legend title
```

Animation of the top 10 observed species through time. Observations are represented as a cumulative sum.

```{r fig.height=6, fig.width=8}

state_birds %>% 
  arrange(date) %>% 
  group_by(primary_com_name) %>% 
  mutate(cumulative_count = cumsum(how_many)) %>% 
  filter(primary_com_name %in% top10) %>% 
  ggplot(aes(date, cumulative_count)) +
  geom_point() + geom_line() +
  facet_wrap(~primary_com_name) +
  transition_reveal(date)

```

## Mapping With Leaflet

Map of all the bird observations in this state for the study duration. The radius of each point indicates the relative number of individuals observed for a given species.

```{r}

leaflet(state_birds, width = "100%") %>% 
  # basemap options
  addProviderTiles(providers$CartoDB, group = "CartoDB") %>% 
  addTiles(group = "OSM (default)") %>%
  addCircleMarkers(radius = ~how_many/5, 
                   group = "Markers",
                   label = ~primary_com_name) 
    # addLayersControl(
    #   baseGroups = c("CartoDB", "OSM (default)"),
    #                  overlayGroups = c("Markers"),
    #                  position = "topright")

```

Let's add some color and look at the top observed species for this state.

```{r}

species_obs <- state_birds %>% filter(primary_com_name %in% top10[1:2])

# Continuous palette
# pal <- colorNumeric(palette = "RdBu", domain = species_obs$how_many)

# Discrete palette
pal <- colorFactor("Set1", levels = species_obs$primary_com_name)

leaflet(species_obs, width = "100%") %>% 
        # basemap options
  addProviderTiles(providers$CartoDB, group = "CartoDB") %>% 
  addTiles(group = "OSM (default)") %>%
  addCircleMarkers(radius = ~how_many, 
                   group = "Markers",
                   color = ~pal(primary_com_name),
                   label = ~how_many) %>% 
  addLegend(position = "bottomright",
            pal = pal, values = ~primary_com_name,
            title = "Legend",
            opacity = 1) %>%
    addLayersControl(
      baseGroups = c("CartoDB", "OSM (default)"),
                     overlayGroups = c("Markers"),
                     position = "topright")


## And many more possibilities!
# https://r-charts.com/spatial/interactive-maps-leaflet/
```

