---
title: "Palmer Penguin Data Visualization"
subtitle: "SERC Data Science"
author: "Jaxine Wolfe"
date: "August 19, 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

library(tidyverse)
library(palmerpenguins)
```

## Tables 

```{r}
smry <- penguins %>% 
  group_by(species) %>% 
  summarize(across(where(is.numeric), mean, na.rm = TRUE)) %>% 
  mutate(across(where(is.numeric), round, 5)) %>% 
  select(-year)

knitr::kable(smry) # static table
```

## Data Distributions

Density Plot

geom_density uses a kernel density estimation (KDE) to create a smooth estimate of the probability density function of the continuous variable. It essentially places a small "kernel" (often a normal distribution) at each data point and sums these kernels to create a continuous curve.

```{r}
penguins %>% 
  ggplot(aes(bill_length_mm, fill = species)) +
  geom_density(alpha = 0.3) + geom_rug(aes(col = species)) +
  theme_bw()
```

Histogram

geom_freqpoly works by dividing the continuous variable into bins, similar to a histogram. It then calculates the count of observations within each bin.

```{r}
penguins %>% 
  ggplot(aes(body_mass_g, fill = sex)) +
  geom_histogram(alpha = 0.5) +
  facet_wrap(~species, ncol = 1) +
  theme_bw()
```

<!-- ## Heat Maps -->

```{r}

# penguins %>% 
#   group_by(species, sex) %>% 
#   summarize(how_many = n(),
#             bill_length_mm = mean(bill_length_mm, na.rm = T),
#             flipper_length_mm = mean(flipper_length_mm, na.rm = T))  %>% 
#   ggplot(aes(species, sex, fill = flipper_length_mm)) +
#   geom_tile(color = "white") +
#   coord_equal() 

```

## Change Through Time

Here is a [resource for color palettes](https://r-graph-gallery.com/ggplot2-color.html)

```{r}

#Create a custom color scale
library(RColorBrewer)
myColors <- brewer.pal(3,"Set2")
names(myColors) <- levels(penguins$species)
colScale <- scale_colour_manual(name = "species", values = myColors)

penguins %>%
  group_by(species, year) %>% 
  summarize(n = n()) %>% 
  ggplot(aes(year, n, col = species)) +
  geom_point() + geom_line() +
  xlab("Year") + ylab("Number of Individuals") +
  theme_bw() + 
  scale_color_hue()
  # scale_color_brewer()
  # colScale

```

## Relationships

```{r}
penguins %>% 
  drop_na(sex) %>% 
  ggplot(aes(body_mass_g, flipper_length_mm, col = species)) +
  geom_point(pch = 1, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal()

```


`r citation("palmerpenguins")`