---
title: "Palmer Penguin Data Visualization"
subtitle: "SERC Data Science"
author: "Jaxine Wolfe"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(palmerpenguins)
library(DT)
library(plotly)
library(gganimate)
# library(gifski)
# library(png)

```

## Meet the penguins

The `penguins` data from the [**palmerpenguins**](https://allisonhorst.github.io/palmerpenguins/index.html) package contains size measurements for `r nrow(penguins)` penguins for three species observed on three islands in the Palmer Archipelago, Antarctica.

## Tables with DT

```{r}
smry <- penguins %>% 
  group_by(species, sex, island) %>% 
  summarize(across(where(is.numeric), mean, na.rm = TRUE)) %>% 
  mutate(across(where(is.numeric), round, 5)) %>% 
  select(-year)

DT::datatable(smry, 
          caption = "Summary of Continuous Variables", 
          options = list(autoWidth = TRUE,
                         paging = FALSE,
                         info = FALSE,
                         scrollY = 300),
          rownames = FALSE)
```

## Plotly

### Scatterplot

```{r}
plot_ly(data = penguins, x = ~bill_length_mm, y = ~bill_depth_mm, 
        color = ~species, colors = "Set1", alpha = 0.5)

# add hover text
# text = ~paste("Sex: ", sex, '<br>Year:', year)
```

### Barplots

```{r}
penguins %>% 
  group_by(species) %>% 
  summarize(n = n()) %>% 
       plot_ly(x = ~species, y = ~n, type = "bar",
               color = ~species, colors = "Set1", width = ) %>% 
       layout(title = "This is a figure title",
              xaxis = list(title = "Species"),
              yaxis = list(title = "Number of Individuals"))

```

Lets add errorbars!

```{r}
bill_length_smry <- penguins %>% drop_na(sex, bill_length_mm) %>% 
  group_by(species, sex) %>%
  summarize(bill_length_mm_se = sd(bill_length_mm)/sqrt(length(bill_length_mm)), # this has to be first
            bill_length_mm = mean(bill_length_mm, na.rm = T)) 

plot_ly(bill_length_smry, x = ~species, y = ~bill_length_mm, type = "bar", color = ~sex,
             error_y = list(array = ~bill_length_mm_se, color = "black"))
```


## Gganimate

### Data Relationships

Animating the body mass to flipper length relationship across different species of penguin. 

```{r}
penguins %>% 
          ggplot(aes(body_mass_g, flipper_length_mm, col = species)) +
          geom_point(size = 2) +
           transition_states(species,
                    transition_length = 2,
                    state_length = 1) +
  enter_fade() +
  exit_shrink()

```

### Timeseries

Adding in sex as a categorical variable and animating the relationship of body mass to flipper length through time.

```{r}

anim <- penguins %>% drop_na(sex) %>% 
          ggplot(aes(body_mass_g, flipper_length_mm, col = sex)) +
          geom_point() +
          stat_ellipse(geom = "polygon",
               aes(fill = sex), 
               alpha = 0.25) + # can also adjust confidence interval with "level"
          facet_wrap(~species) +
          transition_time(year) +
          labs(title = "Year: {frame_time}")

animate(anim, nframes = 3, fps = 0.5)

# animate(anim, nframes = 3, fps = 0.5, renderer = gifski_renderer("figures/mass_flipper_relationship.gif")) 

```

Let's look at another type of animated timeseries.

```{r}
penguins %>% 
  group_by(species, year) %>% 
  summarize(n = n()) %>% 
  ggplot(aes(year, n, col = species)) +
  geom_point() + geom_line() +
  transition_reveal(year) 

anim_save("figures/population_change.gif")

```
